---
title: "Human and Property Cost Analysis from Storm Data"
author: "David C Street"
date: "22 March 2015"
output:
  html_document:
    toc: yes
---

#Synopsis
I have analysed storm data from NOAA's storm database to find the most costly and dangerous storm events. The results are unambigous, a relatively small number of weather event types cause most damage and human casualties. The categories of the data are slightly misleading though. Additional data cleanup is recommended to get more useful results.

#Data Processing
We first need to load relevant libraries and load the storm data into R. You'll also see that I have trimmed the data set in places to make working with it more efficient.

```{r, message=F, results='hide'}
#script to process storm data and do some analysis on it
library(ggplot2)
library(data.table)
library(dplyr)

#file and data processing
originalfile <- 'StormData.csv'
cleanfile <- 'stormdataclean.csv'
#make a nice CSV file and save it
if (!file.exists(cleanfile)){
    rawstormdata <- read.csv(originalfile)
	rawstormdata <- tbl_df(rawstormdata)
	#get rid of some unused columns to speed up operations
	rawstormdata <- select(rawstormdata, -REMARKS, -STATE__,
		-(BGN_RANGE:END_LOCATI), -(CROPDMG:ZONENAMES))
	write.csv(rawstormdata, cleanfile)
}

#load data into local data table, make a couple changes
if (!exists('stormdata')){
	stormdata <- tbl_dt(fread(cleanfile, stringsAsFactors=T))
}
```

Next we take that data a wrangle it into some convenient grouped sets. This allows to easily extract the summary information we need.

First we'll do this for the human casualties of the storm events:

```{r}
#human costs
stormdata_byevent <- group_by(stormdata, EVTYPE)
stormdata_byevent_humansums <- tbl_df(summarise(stormdata_byevent, 
    fatalities=sum(FATALITIES, na.rm=T), injuries = sum(INJURIES, na.rm=T),
	casualties=sum(FATALITIES, INJURIES, na.rm=T)))
stormdata_byevent_humansums <- filter(stormdata_byevent_humansums, casualties>0)
stormdata_byevent_humansums <- arrange(stormdata_byevent_humansums,
	desc(casualties))
stormdata_byevent_humansumstop5 <- head(stormdata_byevent_humansums, n=5)
```

And next for the property damage cause by the storms. Note that is this case, the manipulation of the data is more involved as we have to combine two columns, one numeric and one character, to get the actual numeric property damage. Other than that the process is identical to the compilation of the human casualty data.

```{r}
#property costs    
stormdata_byeventpdmg <- group_by(stormdata, EVTYPE, PROPDMGEXP)
stormdata_byevent_propsums <- tbl_df(summarise(stormdata_byeventpdmg, 
	damages=sum(PROPDMG, na.rm=T)))
stormdata_byevent_propsums <- filter(stormdata_byevent_propsums, damages>0)
#one last processing step to get apply PROPDMGEXP column to sums
#create a new column of data that translates EXP tag into a multiplier
allexp <- stormdata_byevent_propsums[,'PROPDMGEXP']
multy <- numeric()
for (i in 1:dim(allexp)[1]){
	crnt_str <- as.character(allexp[i,])
	if (crnt_str=='M' | crnt_str=='m'){
		multy <- c(multy, 1000000.0)
	} else if (crnt_str=='B' | crnt_str=='b'){
		multy <- c(multy, 1000000000.0)
	} else {
		#assume if it is K, k or blank, the multiplier in 1,000
		multy <- c(multy, 1000.0)
	}
}
#use that multiplier to create a columns that is the actual numerical damage
stormdata_byevent_propsums <- cbind(stormdata_byevent_propsums, multy=multy)
stormdata_byevent_propsums <- mutate(stormdata_byevent_propsums, 
	propdmgtotal=damages*multy)
#group by one more time to consolidate the event types in to one row
stormdata_byeventpdmg <- group_by(stormdata_byevent_propsums, EVTYPE)
stormdata_byevent_propsums <- summarise(stormdata_byeventpdmg,
	damages= sum(propdmgtotal))
stormdata_byevent_propsums <- arrange(stormdata_byevent_propsums,
	desc(damages))
stormdata_byevent_propsumstop5 <- head(stormdata_byevent_propsums, n=5)
```

#Results
To get a sense of scale, let's first look at simple graphs of the human and property costs of the most damaging event types:

In terms of human casualties:
```{r}
humanbartop5 <- ggplot(head(stormdata_byevent_humansumstop5))
humanbartop5 <- humanbartop5 + geom_bar(aes(x=EVTYPE, y=casualties),
    stat='identity')
humanbartop5 <- humanbartop5 + labs(title='5 Most Dangerous Events',
	x='Event', y='Casualties')
humanbartop5
```

In terms of property damage:
```{r}
propbartop5 <- ggplot(head(stormdata_byevent_propsumstop5))
propbartop5 <- propbartop5 + geom_bar(aes(x=EVTYPE, y=damages),
    stat='identity')
propbartop5 <- propbartop5 + labs(title='5 Most Costly Events',
	x='Event', y='Property Damage (USD)')
propbartop5
```

Let's also take a look at tables of the most damaging events:

Human casualties:
```{r, results='markup'}
head(stormdata_byevent_humansums, n=15)
```

Property damage:
```{r, results='markup'}
head(stormdata_byevent_propsums, n=15)
```

Note that there are many different kinds of floods on the property damage table. This is one place where combining categories or coming up with a hierarchy of event types might help us draw more meaning from the data.

#Conclusion
With this data analysis we have a good idea of what event types are most damaging to human health and property. With additional data cleanup, we could put a finer point on this.