---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
    toc: yes
---


## Loading and preprocessing the data
I start by loading the data and using `dplyr` functions to organize it in a way that will make following calculations and plots easier to do. I also use `lubridate` to handle the date and time info a little more gracefully. All plots in this report were made with `ggplot2`, so that library is loaded here as well.

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)

#load data to a data frame
dfraw <- read.csv("./activity.csv")

#create a tbl_df to manipulate with dplyr
tdfraw <- tbl_df(dfraw)
rm(dfraw)

#mutate data as needed to make it easier to work with
tdf1 <- transmute(tdfraw, date=ymd(date), interval, steps)

createweekendfactor <- function(wdayints){
    #create empty container for char vector
	charout <- character(length(wdayints))
	#get weekend and weekday values
	weekdayind <- which(wdayints>1 & wdayints<7)
	weekendind <- which(wdayints==1 | wdayints==7)
	#assign values
	charout[weekdayind] <- 'weekday'
	charout[weekendind] <- 'weekend'
	
	return(as.factor(charout))
}

tdf <- mutate(tdf1, 
	dayofweek = wday(date), 
	partofweek = createweekendfactor(dayofweek))
rm(tdf1)

#create some useful grouped data sets
tdf_bydate <- group_by(tdf, date)
tdf_byday <- group_by(tdf, dayofweek)
tdf_bydayint <- group_by(tdf, dayofweek, interval)
tdf_bypartint <- group_by(tdf, partofweek, interval)
tdf_byint <- group_by(tdf, interval)

#create mean and median data from grouped data
ms_bydate <- summarise(tdf_bydate,
	meanstepbyint=mean(steps, na.rm=T),
	sumstepbydate=sum(steps, na.rm=T))
	 
ms_byint <- summarise(tdf_byint,
	meanstepbyint=mean(steps, na.rm=T),
	sumstepbyint=sum(steps, na.rm=T))
	
m_bydayint <- summarise(tdf_bydayint,
	meanstepbyint = mean(steps, na.rm=T))
```


## What is mean total number of steps taken per day?
Start with s simple histogram of the sum of steps taken each day, NA values have been ignored in these calculations.

```{r}
histsumbydate <- ggplot(data=ms_bydate, aes(x=sumstepbydate))
histsumbydate + geom_histogram()
```

Next for the mean and median calculations. And let's look at the head of the summarisedata as well.

```{r}
meansumperday <- mean(ms_bydate$sumstepbydate, na.rm=T)
medsumperday <- median(ms_bydate$sumstepbydate, na.rm=T)
ms_bydate
```

The mean is `r meansumperday` and the median is `r medsumperday`. Again, NA values have been removed for these calculations.

## What is the average daily activity pattern?
Start with a line graph to get an idea of what the data looks like for each interval, on average across all the days. NA values have been removed in these calculations.

```{r}
linebyint <- ggplot(data=ms_byint)
linebyint + geom_line(aes(x=interval, y=meanstepbyint))
```

Let's also take a look at a sample of the summarised data and calculate the max value.

```{r}
maxstepint <- max(ms_byint$meanstepbyint, na.rm=T)
ms_byint
```

And here is the data arranged in descending order of average steps per interval.

```{r}
arrange(ms_byint, desc(meanstepbyint))
```

The max value is `r maxstepint`. NA values obviously being ignored.


## Imputing missing values
To fill in the NA values we've been ignoring, let's use the average for that day of the week and that interval. The following code accomplishes that:

```{r}
tdfnaind <- which(is.na(tdf$steps))
num_na <- length(tdfnaind)

#create a copy to add imputed values to
imptdf <- tdf

for (ind in tdfnaind) {
    #find the average for this day of week and this interval
	crnt_day <- tdf[[ind, 'dayofweek']]
	crnt_int <- tdf[[ind, 'interval']]
	crnt_dayintavg <- filter(m_bydayint, 
		dayofweek==crnt_day, interval==crnt_int)[[1, 'meanstepbyint']]
		
	#add value to new data frame
	imptdf[ind, 'steps'] <- crnt_dayintavg
}
```

And let's also recreate the same useful grouped data sets that we did for the previous data frame.
```{r}
#create some useful grouped data sets with imputed data
imptdf_bydate <- group_by(imptdf, date)
imptdf_bydayint <- group_by(imptdf, dayofweek, interval)
imptdf_bypartint <- group_by(imptdf, partofweek, interval)
imptdf_byint <- group_by(imptdf, interval)

#create mean and median data from grouped data
msimp_bydate <- summarise(imptdf_bydate,
    meanstepbyint=mean(steps, na.rm=T),
	sumstepbydate=sum(steps, na.rm=T))
	 
msimp_byint <- summarise(imptdf_byint,
	meanstepbyint=mean(steps, na.rm=T),
	sumstepbyint=sum(steps, na.rm=T))
	
mimp_bydayint <- summarise(imptdf_bydayint,
	meanstepbyint = mean(steps, na.rm=T))
	
mimp_bypartint <- summarise(imptdf_bypartint,
	meanstepbyint = mean(steps, na.rm=T))
```

Finally let's look at the data and calculate some useful summary statistics.

```{r}
msimp_bydate

histimpsumbydate <- ggplot(data=msimp_bydate, aes(x=sumstepbydate))
histimpsumbydate + geom_histogram()

meanimpsumperday <- mean(msimp_bydate$sumstepbydate)
medimpsumperday <- median(msimp_bydate$sumstepbydate)
maximpstepint <- max(msimp_byint$meanstepbyint)
```

With a new mean of `r meanimpsumperday` and median `r medimpsumperday`, compared with a mean of `r meansumperday` and median of `r medsumperday`, imputing has clearly had the effect of adding steps, as we would expect since NA's were essentially treated as zeros before. Judging by the histogram it has also obviously reduced the number of very low step days, most of those being previously days without data.

## Are there differences in activity patterns between weekdays and weekends?
Let's take a quick look at differences between days graphically. First with a plot of weekdays vs weekends:

```{r}
linebyintpart <- ggplot(data=mimp_bypartint, aes(x=interval, y=meanstepbyint))
linebyintpart + geom_line() + facet_wrap(~partofweek)
```

And also for a little more granularity by days of the week. Sunday is 1, Monday is 2, etc...

```{r}
scatbyintday <- ggplot(data=m_bydayint, aes(x=interval, y=meanstepbyint))
scatbyintday + geom_line() + facet_wrap(~dayofweek)
```

There does seem to be fairly large differences based on the time of the week judging from these plots.