---
title: "Tooth growth data analysis"
author: "David C Street"
date: "22 March 2015"
output:
  pdf_document: default
---

#Summary
This report presents an analysis of the `ToothGrowth` data set in R. It includes an exploratory data analysis and statistical inference data analysis.

#Data Processing
We need to first load appropriate libraries and the data. I have also performed a couple small data transformations to make data analysis easier. The last line calculates summary statistics on the grouped data. Here is the R code to do that:

```{r, message=F}
library(ggplot2)
library(dplyr)

data(ToothGrowth)

#make a couple tbl_df's to facilitate data analysis
tg <- tbl_df(ToothGrowth)
tg_dose <- mutate(tg, dosefact=as.factor(dose))

tg_bysuppdose <- group_by(tg, supp, dose)
tg_bysuppdose_stat <- summarise(tg_bysuppdose, 
    meanlen=mean(len), sdlen=sd(len))
```

#Exploratory Data Analysis
Let's first look at the summary statistics of the data:
```{r, echo=F}
tg_bysuppdose_stat
```

And let's look at a violin plot of the data to get a visual idea of the shape of the data:

```{r}
vplottitle <- 'Tooth growth by supplement (OJ=Orange Juice, VC=Ascorbic Acid)'
vplot <- ggplot(tg_dose, aes(y=len, x=dosefact))
vplot <- vplot + geom_violin() + facet_wrap(~supp)
vplot <- vplot + labs(title=vplottitle, 
    x='vitamin C dose (mg)', y='tooth length')
vplot
```

It seems clear that there is a significant effect on tooth growth with both supplementation techniques based on this quick look. That is, increasing the dose increases tooth growth. It is less if clear if there is a significant difference between the methods of supplementation, orange juice (OJ) versus ascorbic acid (VC).

Next we'll see what statistical inference shows.

#Statistical Inference Data Analysis
##Dosing
First, considering only the dose and not the supplementation method, let's see if we can find a statistically significant difference between a 0.5 mg dose and a 2.0 mg dose.

```{r}
tg_bydose <- group_by(tg, dose)
tg_bydosemean <- summarise(tg_bydose, meanlen=mean(len))
tg_bydosemean
```

The mean of the tooth length is higher for each dose increase. So let's test the hypotheses that each step up increases tooth growth.

Those tests set up in R:
```{r}
dose1 <- filter(ToothGrowth, dose==0.5)[,'len']
dose2 <- filter(ToothGrowth, dose==1.0)[,'len']
t1 <- t.test(dose1, dose2, alternative='greater')

dose1 <- filter(ToothGrowth, dose==1.0)[,'len']
dose2 <- filter(ToothGrowth, dose==2.0)[,'len']
t2 <- t.test(dose1, dose2, alternative='greater')

dose1 <- filter(ToothGrowth, dose==0.5)[,'len']
dose2 <- filter(ToothGrowth, dose==2.0)[,'len']
t3 <- t.test(dose1, dose2, alternative='greater')
```

Based on the results of each one of these t-tests shows that we can reject the null hypothesis with 95% confidence in each case. In other words each increase in dose shows a statistically significant tooth growth.

##Supplementation method
Next let's see if we can find a statistically significant differences between supplement methods.

```{r}
tg_bysupp <- group_by(tg, supp)
tg_bysuppmean <- summarise(tg_bysupp, meanlen=mean(len))
tg_bysuppmean
```

OJ supplementation has a higher mean so let's test the hypothesis that is increases tooth growth more than VC (ascorbic acid).

That test set up in R:
```{r}
supp1 <- filter(ToothGrowth, supp=='VC')[,'len']
supp2 <- filter(ToothGrowth, supp=='OJ')[,'len']
t4 <- t.test(supp1, supp2, alternative='greater')
```

Based on the t-test results, we can reject the null hypothesis with 95% confidence. In other words, the orange juice shows a statistically significant increase in tooth growth versus the ascorbic acid.

#Appendices
##Results of t-tests
The detailed results of the four t-tests performed:
```{r}
t1
t2
t3
t4
```

